---
title: "Generic Analysis"
date: '2020-09-11'
author: 'David McGaughey'
output: 
  html_notebook:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
---

```{r, message=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(edgeR)
library(data.table)
library(tximport)
library(DESeq2)
library(ggrepel)
library(IHW)
library(ComplexHeatmap)
library(annotables)
# Parallel
library(BiocParallel)
register(MulticoreParam(12))

working_dir <- '/Volumes/data/projects/nei/keep/going'
files <- list.files(path=working_dir,recursive=TRUE,pattern='quant.sf', full.names = TRUE)
```

```{r "setup", include=FALSE}
require("knitr")
opts_knit$set(root.dir = working_dir)
```

# start building anno to go from transcript to gene name to common gene name
```{r, message=FALSE, message=FALSE, warning=FALSE}
anno <- fread(files[1]) # any file will work
anno_tximport <- anno %>% rowwise() %>% 
  mutate(transcript_id = Name,
         gene_name = str_split(transcript_id, '\\|')[[1]][6])

```


# actually merge tx specific counts to gene level
```{r, message=FALSE, warning=FALSE}
txi <- tximport(files, type = "salmon", tx2gene = anno_tximport %>% select(Name, gene_name))
txi.deseq2 <- data.frame(txi$counts)
```

# Build Metadata
```{r, message=FALSE, message=FALSE, warning=FALSE}
meta <- files %>% 
  #gsub('/quant.sf','',.) %>% 
  as_tibble() %>% 
  rowwise() %>% 
  mutate(file = value,
         value = gsub('/quant.sf', '', file), 
         value = (str_split(value, '/')[[1]] %>% rev())[1],
         sample = str_extract(value, 'AMDCD|D3C|TJP1'),
         dose = str_extract(value, '^\\d+') %>% as.numeric(),
         dose = case_when(is.na(dose) ~ 0, TRUE ~ dose),
         replicate = str_extract(value, '\\d$') %>% as.factor(),
         drug = case_when(dose == 0 ~ 'No',
                          TRUE ~ 'Yes'),
         ID = paste(sample, replicate, drug, dose, sep = '_')) %>% 
  dplyr::select(-value)
meta %>% DT::datatable()
```

# Calculate TPM scores by type, correct for library size
```{r}
#library(edgeR)
txi_TPM <- tximport(meta$file,
                    type = "salmon",
                    tx2gene = anno_tximport %>% select(Name, gene_name),
                    countsFromAbundance = "lengthScaledTPM")
tpms <- as.data.frame(txi_TPM$counts)
colnames(tpms) <- meta$file
norm <- edgeR::DGEList(tpms)
norm <- edgeR::calcNormFactors(norm)
norm_counts <- norm$counts
#extract scaling factor for each sample and  multiply
correction <- norm$samples %>% data.frame() %>% .[['norm.factors']]
lsTPM_librarySize <- norm_counts %*% diag(correction)
colnames(lsTPM_librarySize) <- meta$file
lsTPM <- lsTPM_librarySize %>%
  as_tibble(rownames = 'Gene') %>% 
  gather('file', 'lsTPM', -Gene) %>% 
  left_join(., meta, by = 'file') %>% 
  left_join(grch38 %>% dplyr::select(Gene=symbol, description))

```


# DESeq2

```{r, message=FALSE, message=FALSE, warning=FALSE}
dds <- DESeqDataSetFromTximport(txi, meta, ~sample + replicate + drug)
DESeq2Table <- DESeq(dds, parallel = TRUE)
vst <- varianceStabilizingTransformation(DESeq2Table)
colnames(vst) <- paste(colData(dds)$sample, colData(dds)$replicate, colData(dds)$drug, colData(dds)$dose, sep='_')
```

# PCA clustering with 1000 most variable genes
```{r, message=FALSE, message=FALSE, warning=FALSE}
#PCA with 1000 most variable genes
ntop = 1000
Pvars <- rowVars(assay(vst))
select <- order(Pvars, decreasing = TRUE)[seq_len(min(ntop, length(Pvars)))]
PCA <- prcomp(t(assay(vst)[select, ]), scale = F)
percentVar <- round(100*PCA$sdev^2/sum(PCA$sdev^2),1)
dataGG = data.frame(PC1 = PCA$x[,1], PC2 = PCA$x[,2], 
                    PC3 = PCA$x[,3], PC4 = PCA$x[,4], 
                    drug = colData(vst)$drug,
                    sample = colData(vst)$sample,
                    dose = colData(vst)$dose)
dataGG %>% 
  mutate(dose = as.factor(dose)) %>% 
  ggplot(aes(PC1, PC2, color=dose, shape=sample)) +
  geom_point(size=3) + 
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + ggtitle("PCA") + theme_minimal() + ggsci::scale_colour_aaas() 


dataGG %>% 
  mutate(dose = as.factor(dose)) %>% 
  ggplot(aes(PC3, PC4, color=dose, shape=sample)) +
  geom_point(size=3) + 
  xlab(paste0("PC3: ",percentVar[3],"% variance")) +
  ylab(paste0("PC4: ",percentVar[4],"% variance")) + ggtitle("PCA") + theme_minimal() + ggsci::scale_colour_aaas() 

```

# sample distances
```{r, message=FALSE, message=FALSE, warning=FALSE}
sampleDists <- dist(t(assay(vst)))
sampleDistMatrix <- as.matrix(sampleDists) %>% scale(., center= FALSE)
Heatmap(sampleDistMatrix, col = circlize::colorRamp2(seq(0,1,0.1), colors = viridis::viridis(n = 11)))
```

```{r, message=FALSE, message=FALSE, warning=FALSE}
process <- function(txi, metadata, contrast, design){
  dds <- DESeqDataSetFromTximport(txi, metadata, design)
  print(design(dds))
  DESeq2Table <- DESeq(dds, parallel = TRUE)
  DESeq2Res <- results(DESeq2Table, 
                       contrast=contrast, 
                       filterFun=ihw)
  
  # hist of pvalues
  print(hist(DESeq2Res$pvalue, col = "lavender", main = paste(contrast[2], ' vs ', contrast[3], ', all values', sep =''), xlab = "p-values"))
  
  # Filter by padj and padjust to get rid of NA
  DESeq2Res <- DESeq2Res[ !is.na(DESeq2Res$padj), ]
  DESeq2Res <- DESeq2Res[ !is.na(DESeq2Res$pvalue), ]
  
  # Only keep genes with baseMean > 0.5
  DESeq2Res <- DESeq2Res[ (DESeq2Res$baseMean > 0.5), ]
  
  print(hist(DESeq2Res$pvalue, col = "royalblue",  main = paste(contrast[2], ' vs ', contrast[3], ', remove NA and low counts', sep = ''), xlab = "p-values"))
  
  # MA plot
  print(DESeq2::plotMA(DESeq2Res))
  
  # output data
  stats <- DESeq2Res %>% as_tibble(rownames = 'Ensembl Gene')
  gene_counts <- counts(DESeq2Table, normalized=TRUE) %>% as_tibble(rownames = 'Ensembl Gene')
  
  
  # add Gene common name
  stats <- left_join(stats, grch38 %>% dplyr::select(`Ensembl Gene` = symbol, description))
  
  # order by pvalue and add FC column
  stats <- stats %>% arrange(pvalue) %>% mutate(FoldChange = 2^log2FoldChange) %>% unique()
  
  return(list(Stats = stats, Count = gene_counts))
}
```

```{r}
diff_testing <- suppressWarnings(process(txi, meta, c('drug','Yes','No'), ~sample + replicate + drug))
colnames(diff_testing$Count)[2:ncol(diff_testing$Count)] <- dds@colData$ID
```

# Volcano Plot code
```{r, set.width=3}
volcano_maker <- function(df, 
                          title = '',
                          padj_cutoff_for_labels = 1e-10, 
                          abs_log2fc_cutoff_for_label = 1,
                          coord_cartesian = c(-3, 3),
                          x_intercept_dotted = c(-2,2)){
  df$Class <- ifelse(df$padj < 0.01, "FDR < 0.01", "Not Significant")
  df$Class[df$padj < 0.01 & abs(df$log2FoldChange) > 1] <- "FDR < 0.01 & log2FC > 1"
  df$Class <- factor(df$Class,levels=c("Not Significant","FDR < 0.01", "FDR < 0.01 & log2FC > 1"))
  plot <- ggplot(data=df,aes(x=log2FoldChange,y=-log10(pvalue))) + 
    geom_point(aes(colour = Class), size=0.7) +
    scale_colour_manual(values=c("gray","pink","darkred")) + 
    geom_label_repel(data=subset(df,abs(log2FoldChange) > abs_log2fc_cutoff_for_label & padj< padj_cutoff_for_labels), 
                    aes(label=`Ensembl Gene`)) +
    geom_vline(aes(xintercept=x_intercept_dotted[2]),linetype="dotted") +
    scale_x_continuous(breaks=c(seq(-100,100,1))) +
    coord_cartesian(xlim = c(coord_cartesian[1], coord_cartesian[2])) +
    geom_vline(aes(xintercept=x_intercept_dotted[1]),linetype="dotted") +
    ggtitle(title) +
    cowplot::theme_cowplot()
  return(plot)}

scatter_maker <- function(lsTPM, genes = c('RHO'), 
                          group_on_col, 
                          color_on_col = NULL,
                          shape_on_col = NULL){
  lsTPM %>% 
    filter(Gene %in% genes) %>% 
    ggplot(aes(x=!!enquo(group_on_col), 
               y = log2(lsTPM+1), 
               group = !!enquo(group_on_col),
               color = !!enquo(color_on_col),
               shape = !!enquo(shape_on_col))) + 
    geom_boxplot(width = 0.5) + 
    geom_point(size = 2) +
    cowplot::theme_cowplot() + 
    facet_wrap(~Gene)
}


```
# Volcano Plot
```{r}
volcano_maker(diff_testing$Stats)
```

# Box plot
```{r, fig.height = 12, fig.width = 10}
scatter_maker(lsTPM, 
              diff_testing$Stats %>% 
                filter(abs(log2FoldChange) > 1, padj < 1e-10) %>% 
                pull(`Ensembl Gene`), 
              drug, 
              sample, 
              as.factor(dose))
```

# GO Enrichment
## Table
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
all_genes <- bitr(diff_testing$Stats %>% pull(`Ensembl Gene`), fromType = "SYMBOL",
        toType = c("ENTREZID", "ENSEMBL"),
        OrgDb = org.Hs.eg.db)

diff_genes <- bitr(diff_testing$Stats %>% filter(padj < 0.01) %>% pull(`Ensembl Gene`), fromType = "SYMBOL",
        toType = c("ENTREZID", "ENSEMBL"),
        OrgDb = org.Hs.eg.db)

ego <- enrichGO(gene          = diff_genes$ENTREZID,
                universe      = all_genes$ENTREZID,
                OrgDb         = org.Hs.eg.db,
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
        readable      = TRUE)

ego %>% as_tibble() %>% DT::datatable()

```
## Dot Plot
```{r}
dotplot(ego)
```
# Kegg Pathway Enrichment
## Table
```{r}
kk <- enrichKEGG(gene         = diff_genes$ENTREZID,
                 organism     = 'hsa',
                 pvalueCutoff = 0.05)
kk %>% as_tibble() %>% DT::datatable()
```

## Dotplot
```{r}
dotplot(kk)
```

```{r}
devtools::session_info()
```
