---
title: "Analysis Template"
author: "David McGaughey"
date: "`r Sys.Date()`"
output:
 html_notebook:
  theme: flatly
  toc: true
  toc_float: true
  code_folding: hide
 html_document:
  theme: flatly
  toc: true
  toc_float: true
  code_folding: hide
---

/Users/mcgaugheyd/git/OGVFB_RNAseq

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,  warning = FALSE,
  collapse = TRUE,
  fig.width = 12, fig.height = 8,
  comment = "#>",
        dpi=300
    )
```



```{r, message=FALSE, message=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(biomaRt)
library(tximport)
library(DESeq2)
library(stringr)
library(ggrepel)
library(apeglm)
library(clusterProfiler)
library(matrixStats)
library(mcHelpeRs) # devtools::install_github("davemcg/mcHelpeRs")
#library(ReactomePA)
library(PCAtools)
# Parallel
library(BiocParallel)
register(MulticoreParam(12))
working_dir <- '~/data/bharti/PROJECT/salmon_quant/'
files <- list.files(path=working_dir,recursive=TRUE,pattern='quant.sf', full.names = TRUE)
```

# Extract gene annotation
```{r, message=FALSE, message=FALSE, warning=FALSE}
anno <- fread(files[1]) 
anno$Gene <- sapply(anno$Name,function(x) strsplit(x,'\\|')[[1]][6])
anno_tximport <- anno %>% 
  dplyr::select(target_id = Name, Gene)
```

# Metadata
```{r}
meta <- gsub(pattern = '\\/Users\\/mcgaugheyd\\/data\\/bharti\\/PROJECT\\/salmon_quant\\/\\/','',files) %>% gsub('\\/quant.sf','',.) %>% enframe()
meta$salmon_path <- files

scientist_meta <- readxl::read_xlsx("RNAseq Sample description 2026.xlsx")
meta$treatment <- str_extract(string = meta$value, pattern = "C[CI]HS")
meta <- meta %>% dplyr::rename(sample = value) %>% 
  left_join(connie_meta,  by = c("sample" = "Sample name"))
```

# Import quant at gene level
```{r, message=FALSE, message=FALSE, warning=FALSE}
txi <- tximport(meta$salmon_path, type = "salmon", tx2gene = anno_tximport)
```



# Sex Guess

```{r, fig.height=3, fig.width=3}
load('~/data/metamoRph_models/sex.Rdata')
gcounts <- txi$counts

conv_table <- biomaRt::select(org.Hs.eg.db::org.Hs.eg.db, keys = row.names(gcounts), columns = c("ENSEMBL", "GENENAME", "MAP", "GENETYPE"), keytype = "SYMBOL")
row.names(gcounts) <- row.names(gcounts) %>% 
  enframe() %>% 
  left_join(conv_table %>% dplyr::select(SYMBOL,ENSEMBL) %>% filter(!duplicated(ENSEMBL), !duplicated(SYMBOL)), 
            by = c("value" = "SYMBOL")) %>% pull(ENSEMBL)

# remove NA gene rows (where gene symbol <-> ensembl join didn't work because of multi-matches)
projected_data_outside <- metamoRph::metamoRph(gcounts[!is.na(row.names(gcounts)),], gtex_sex$PCA$rotation, gtex_sex$center_scale)
label_guesses_outside <- metamoRph::model_apply(sex_model, projected_data_outside)

meta$sexGuess <- label_guesses_outside$predict

meta$sexGuess
```

# Tissue Guess

```{r}
load('~/data/metamoRph_models/eiad_ocular_tissue.02.Rdata')

projected_data_outside <- metamoRph::metamoRph(gcounts[!is.na(row.names(gcounts)),], eiad_pca$PCA$rotation, eiad_pca$center_scale)
label_guesses_outside <- metamoRph::model_apply(tissue_model, projected_data_outside)

meta$Tissue_Predict <- label_guesses_outside$predict
meta$Tissue_Score <- label_guesses_outside$max_score

meta$Tissue_Predict
```


# Deseq2 time
Batch is a covariate. Testing against Description. 

```{r}
de <- mcHelpeRs::run_deseq2_analysis(input_data = txi, 
                                     sample_metadata = meta, 
                                     design_formula = "~Description+Batch", use_lrt = FALSE)
```


## Biplot

```{r, fig.width=10, fig.height=5, dpi = 300}
biplot(de$pca_results, 
       colby = 'Description', 
       x = 'PC1',
       y='PC2', 
       lab = NULL, 
       showLoadings = TRUE, 
       shape = 'treatment', 
       legendPosition = 'right') + 
  scale_color_manual(values = pals::okabe() |> unname()) + 
  ggtitle("All Samples")

biplot(de$pca_results, 
       colby = 'Description', 
       x = 'PC3',
       y='PC4', 
       lab = NULL, 
       showLoadings = TRUE, 
       shape = 'treatment', 
       legendPosition = 'right') + 
  scale_color_manual(values = pals::okabe() |> unname()) + 
  ggtitle("All Samples")
```


# Diff Testing

Pull diff testing with a custom contrast.
``` {r, echo=FALSE,warning=FALSE,message=FALSE}
#########
# DE time
#########

comparisonA <- 'CIHS'
comparisonB <- 'CCHS'
res <- results(de$dds, contrast = c('Description',
                                    comparisonA,
                                    comparisonB)) #%>% as_tibble(rownames = 'Gene') %>% arrange(padj, -abs(log2FoldChange))

#plotMA(res)
#res$pvalue %>% hist() %>% plot()
```

## Results
Positive is genes that are more expressed in `CIHS` vs `CCHS`
```{r}
res %>%  
  as_tibble(rownames = 'Gene') %>% 
  mutate(Comparison = paste0(comparisonA,
                             " vs ",
                             comparisonB)) %>% 
  relocate(Comparison) %>% 
  arrange(padj) %>% head(100) %>% DT::datatable()
```

## Dot Plots of Select Genes
```{r, fig.width=18, fig.height=10}
gene_ncounts <- counts(de$dds, normalize = TRUE)
colnames(gene_ncounts) <- de$dds$sample
plot_genes <- bind_rows(subset(res %>% as_tibble(rownames = 'Gene'), padj<0.01 & baseMean > 100 & log2FoldChange > 1) %>% arrange((padj)) %>% head(10), 
                        subset(res %>% as_tibble(rownames = 'Gene'), padj<0.01 & baseMean > 100 & log2FoldChange < -1) %>% arrange((padj)) %>% head(10)) %>% pull(Gene)

gene_ncounts %>% 
  as_tibble(rownames = 'Gene') %>% 
  pivot_longer(-Gene) %>% 
  left_join(colData(de$dds) %>% as_tibble(), by = c('name' = 'sample')) %>% 
  filter(Gene %in% plot_genes, 
         Description %in% c(comparisonA, comparisonB)) %>% 
  ggplot(aes(y=Description, x = log2(value+1))) + 
  geom_boxplot(color = 'black') +
  geom_point() + 
  xlab("log2(norm count)") +
  facet_wrap(~Gene, scales = 'free_x') +
  cowplot::theme_cowplot() 
```


## Volcano Plot

```{r}
mcHelpeRs::volcano_maker(df = res %>% as_tibble(rownames = 'Gene'), 
                         pos_label = "CIHS",
                         neg_label = "CCHS") + 
  coord_cartesian(xlim = c(-7,7))
```


# Enrichment Analysis (GO and Pathway)

## GO Enrichment 

```{r}
# remove low count genes (<1 count per sample)
lc_genes <- de$raw_counts[de$raw_counts %>% rowSums() < ncol(de$raw_counts),] %>% row.names()
resFILTER <- res %>% 
  as_tibble(rownames = 'Gene') %>% 
  filter(!Gene %in% lc_genes)

msig <- mcHelpeRs::prepare_msigdb_sets()

de$ego <- mcHelpeRs::run_ora(universe = resFILTER %>% pull(Gene),
                             diff_genes = resFILTER %>% filter(abs(log2FoldChange) > 1, padj < 0.05),
                             term2name = msig$descriptions, 
                             term2gene = bind_rows(msig$cp, msig$go))

```

<!-- ```{r, fig.height=12, fig.width=50} -->
<!-- de$ego@result %>%  -->
<!--   filter(qvalue < 0.05) %>%  -->
<!--   separate(GeneRatio,sep = '/', into = c("GeneNumerator","GeneDenominator")) %>%  -->
<!--   mutate(GeneCount = as.integer(GeneNumerator), -->
<!--          GeneDenominator = as.integer(GeneDenominator), -->
<!--          GeneRatio = GeneCount/GeneDenominator) %>%  -->
<!--   mutate(Description = forcats::fct_reorder(Description, GeneRatio)) %>%  -->
<!--   mutate(significance = case_when(qvalue < 0.001 ~ "q<0.001", -->
<!--                                   qvalue < 0.01 ~ "q<0.01", -->
<!--                                   qvalue < 0.05 ~ "q<0.05", -->
<!--                                   qvalue < 0.1 ~ "q<0.1", -->
<!--                                   TRUE ~ "q>0.1")) %>%  -->
<!--   ggplot(aes(x=GeneRatio,y=Description,color = significance, size = GeneCount)) + -->
<!--   geom_point() + -->
<!--   cowplot::theme_cowplot() + -->
<!--   scale_color_manual(values = c(viridis::viridis(n=10)[c(10,8,6,4,1)]))  -->
<!-- ``` -->


### Table
```{r}
de$ego@result %>% 
  arrange(pvalue) %>% 
  filter(qvalue < 0.05) %>% 
  DT::datatable()
```

## GSEA

```{r}

de$gsea <- mcHelpeRs::run_gsea(res_df =  resFILTER,
                             term2name = msig$descriptions, 
                             term2gene = bind_rows(msig$cp, msig$go), pvalueCutoff = 1)

```

<!-- ```{r, fig.height=12, fig.width=50} -->
<!-- de$ego@result %>%  -->
<!--   filter(qvalue < 0.05) %>%  -->
<!--   separate(GeneRatio,sep = '/', into = c("GeneNumerator","GeneDenominator")) %>%  -->
<!--   mutate(GeneCount = as.integer(GeneNumerator), -->
<!--          GeneDenominator = as.integer(GeneDenominator), -->
<!--          GeneRatio = GeneCount/GeneDenominator) %>%  -->
<!--   mutate(Description = forcats::fct_reorder(Description, GeneRatio)) %>%  -->
<!--   mutate(significance = case_when(qvalue < 0.001 ~ "q<0.001", -->
<!--                                   qvalue < 0.01 ~ "q<0.01", -->
<!--                                   qvalue < 0.05 ~ "q<0.05", -->
<!--                                   qvalue < 0.1 ~ "q<0.1", -->
<!--                                   TRUE ~ "q>0.1")) %>%  -->
<!--   ggplot(aes(x=GeneRatio,y=Description,color = significance, size = GeneCount)) + -->
<!--   geom_point() + -->
<!--   cowplot::theme_cowplot() + -->
<!--   scale_color_manual(values = c(viridis::viridis(n=10)[c(10,8,6,4,1)]))  -->
<!-- ``` -->


### Table
```{r}
de$gsea@result %>% 
  arrange(pvalue) %>% 
  filter(qvalue < 0.05) %>% 
  DT::datatable()
```

<!-- ### "CNET" Plot -->
<!-- Relationships between related GO terms with shared genes -->
<!-- ```{r, fig.width=6, fig.height=4} -->
<!-- geneList <- eg_diff_genes$log2FoldChange -->
<!-- names(geneList) <- eg_diff_genes$Gene -->
<!-- cnet <- cnetplot(ego, foldChange = geneList) + scale_color_viridis_c(name = 'log2(FoldChange)') -->
<!-- cnet -->
<!-- ``` -->

<!-- ### Enriched GO Term Relationships -->
<!-- Shows the "sets" of GO terms to increase understanding of what the diff gene sets are doing. There are many related GO terms. -->
<!-- ```{r, fig.width=8, fig.height=8} -->
<!-- ego_pair <- pairwise_termsim(ego)  -->
<!-- go_term_relationships <- emapplot(ego_pair,layout="kk") #+ ggtitle("Enriched GO Term Relationships") -->
<!-- go_term_relationships -->
<!-- ``` -->

<!-- ## Wikipathways -->
<!-- ```{r} -->
<!-- # system("wget https://wikipathways-data.wmcloud.org/current/gmt/wikipathways-20230210-gmt-Homo_sapiens.gmt") -->
<!-- wp2gene <- read.gmt('wikipathways-20230210-gmt-Homo_sapiens.gmt') -->
<!-- wp2gene <- wp2gene %>% tidyr::separate(term, c("name","version","wpid","org"), "%") -->
<!-- wpid2gene <- wp2gene %>% dplyr::select(wpid, gene) #TERM2GENE -->
<!-- wpid2name <- wp2gene %>% dplyr::select(wpid, name) #TERM2NAME -->
<!-- ewp <- enricher(eg_diff_genes$ENTREZID, -->
<!--                 TERM2GENE = wpid2gene, -->
<!--                 TERM2NAME = wpid2name, -->
<!--                 pvalueCutoff = 0.1) -->
<!-- ewp_plot <- dotplot(ewp, showCategory=30) + ggtitle("Dotplot for WikiPathways") -->
<!-- ``` -->

<!-- ```{r, fig.width=4, fig.height=3} -->
<!-- ewp_plot -->
<!-- ``` -->
<!-- ### Table -->
<!-- ```{r} -->
<!-- ewp <- setReadable(ewp, OrgDb = org.Hs.eg.db, keyType = 'ENTREZID') -->
<!-- ewp@result %>% DT::datatable() -->
<!-- ``` -->

<!-- ### Heatmaps -->

<!-- #### G protein-coupled receptor activity GO:0004930 -->
<!-- ```{r, fig.width=10, fig.height=8} -->
<!-- relative <- log2(counts(dds, normalize = TRUE) + 1) # add a pseudocount of 1 as log2(0) is infinity -->
<!-- colnames(relative) <- meta$Library -->
<!-- row_means <- rowMeans(relative) -->
<!-- relative <- relative - row_means -->
<!-- #relative_g <-  relative -->

<!-- gene_set<- (gse@result %>% filter(ID == 'GO:0004930') %>% pull(core_enrichment) %>% str_split(., '\\/'))[[1]] -->


<!-- library(ComplexHeatmap) -->
<!-- hm_matrix <- relative[gene_set, meta %>% arrange(Condition, Library) %>% pull(Library)] -->

<!-- ha_column = HeatmapAnnotation(df = data.frame(Condition = meta %>% arrange(Condition, Library) %>% pull(Condition)), -->
<!--                               col = list(Condition = c("GFP+" = viridis::magma(20)[5],  -->
<!--                                                        "ctrl center" = viridis::magma(20)[8], -->
<!--                                                        "ctrl periphery" = viridis::magma(20)[14]))) -->

<!-- Heatmap(hm_matrix,  -->
<!--         cluster_columns = FALSE, -->
<!--         cluster_rows = TRUE,  -->
<!--         name = 'log2(Fold\nChange)',  -->
<!--         top_annotation = ha_column,  -->
<!--         col = circlize::colorRamp2( -->
<!--           seq(-8, 8, length.out = 11),  -->
<!--           RColorBrewer::brewer.pal(11, "RdBu") -->
<!--         )) -->
<!-- ``` -->

<!-- ## KEGG Pathway Enrichment -->
<!-- ```{r, fig.width=3} -->
<!-- kk <- enrichKEGG(gene         = eg_diff_genes$ENTREZID,  -->
<!--                  universe = eg_universe$ENTREZID, -->
<!--                  organism     = 'hsa', -->
<!--                  pvalueCutoff = 0.1) -->
<!-- dotplot(kk) + ggtitle("KEGG Pathway Enrichment")  -->
<!-- ``` -->

<!-- ### Table -->
<!-- ```{r} -->
<!-- kk <- setReadable(kk, OrgDb = org.Hs.eg.db, keyType = 'ENTREZID') -->
<!-- kk@result %>% as_tibble() %>% arrange(p.adjust) %>% filter(p.adjust < 0.01) %>% DT::datatable() -->
<!-- ``` -->

```{r}
# need a gene (symbols) by sample matrix
# that is normalized (at least cpm/log1p)
# or vst
# or something else

# vst (start with this)
dds_filter <-  de$dds[,colData(de$dds)$Description %in% c(comparisonA, comparisonB)] 

norm_name_mat <- assay(dds_filter, 'counts') %>% vst()
colnames(norm_name_mat) <- dds_filter$sample

# filter down to comparisonA and B
# CPM
#norm_name_mat <- de$counts %>% edgeR::cpm() %>% {.+1} %>% log2()
# raw counts (dangerous!)
#norm_name_mat <- de$counts %>% {.+1} %>% log2()
```

```{r}

msig_cp_list <-  split(msig$cp$gene_symbol, msig$cp$gs_name)
msig_go_list <-  split(msig$go$gene_symbol, msig$go$gs_name)

gsva_param <- gsvaParam(
  norm_name_mat,
  c(msig_cp_list, msig_go_list),
  # as there is transformed data
  kcdf = "Gaussian", 
  # ignores gene sets with less than 10 or more than 500 genes
  minSize = 10, maxSize = 500,
)

gsva_res <- gsva(gsva_param, verbose=FALSE)
```

# Diff Expression of Pathways

DESeq2 requires raw (nonnegative) counts. 

limma doesn't really care. 

If you have multiple conditions in your comparison then take some care with the logFC calculations.

```{r, fig.width=16, fig.height=9}
gsva_meta <- colData(dds_filter) %>% as_tibble()
gsva_meta$Description <- relevel(gsva_meta$Description, ref = "Complement competent human serum (CCHS)")
mod <- model.matrix(~factor(gsva_meta$Description))
fit <- lmFit(gsva_res, mod)
fit <- eBayes(fit)

gsva_limma <- gsva_res %>% as_tibble(rownames = 'gene set') %>% 
  left_join(topTable(fit, number = Inf) %>% as_tibble(rownames = 'gene set')) %>% 
  arrange(adj.P.Val, P.Value)

#gsva_limma$adj.P.Val %>% hist()

mcHelpeRs::volcano_maker(gsva_limma %>% mutate(Gene = str_wrap(gsub("_"," ", `gene set`), 20),
                                               baseMean = 1000), 
                         logFC_cutoff = c(-0.5,0.5), 
                         logFC = "logFC", pvalue = "P.Value", padj = "adj.P.Val",
                         pos_label = "CCHS + HERV-K KD",
                         neg_label = "CCHS") +
  xlab("GSVA Pathway Change")
```


# Terms with adjusted Pval < 0.05

I am listing all the genes in the pathway or gene set. This will include genes which are not differentially expressed between your conditions. 
```{r}
gsva_limma %>% 
  filter(adj.P.Val < 0.05) %>% 
  left_join(c(msig_cp_list, msig_go_list) %>% 
              enframe() %>% 
              dplyr::rename(`gene set` = name) %>% 
              unnest(value) %>% group_by(`gene set`) %>% 
              summarise(genes = paste0(value, collapse = ', '))) %>% 
  DT::datatable()
```


# Outputs

The "analysis_00_diff_table.freeze01.csv" file is the diff table from this document with the adjusted logFC

The "analysis_00_diff_table_raw.freeze01.csv" file is the diff table from this document with the raw (unadjusted) logFC

The "analysis_00.rse.geyser.Rdata" is a [geyser](http://github.com/davemcg/geyser) compatible object for interactive expression analysis.

You can use the rse file as follows in R:
```
install.packages('remotes')
remotes::install_github('davemcg/geyser')
load('~/path_to/analysis_00.rse.geyser.Rdata')
library(geyser)
geyser(rse)
```


```{r}
system("mkdir -p ../data")
write_csv(resLFC %>% as_tibble(rownames = 'Gene') %>% arrange(-abs(log2FoldChange)), file = '../data/analysis_00_diff_table.freeze01.csv')
write_csv(res %>% as_tibble(rownames = 'Gene') %>% arrange(-abs(log2FoldChange)), file = '../data/analysis_00_diff_table_raw.freeze01.csv')

library(SummarizedExperiment)
norm_counts <- counts(dds, normalize = TRUE)
colnames(norm_counts) <- dds$sample
rse <- SummarizedExperiment(assays = list(counts = norm_counts,
                                         tpm = tpm), colData = meta)
colnames(rse) <- meta$Library
save(rse, file = '../data/analysis_00.rse.geyser.Rdata')
```



# Session Info
```{r}
devtools::session_info()
```
